---
title: "assignment_4_peter_omasta"
author: "Peter Omasta"
date: "11/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


### INTRODUCTION ###

This is my 4th assignment, where I will be creating a cohesive, well-written, and professionally presented mini-report describing spiny lobster size and abundance for lobsters observed at 5 locations along the Santa Barbara coastline by the SBC LTER.

```{r, include=FALSE}
library(tidyverse)
library(janitor)
library(tidyr)
library(plotly)
library(DT)
```

```{r, include = FALSE}
#We will start by importing the necessary .csv file 

lob_data <- read_csv("lobster_abundance_sbc_lter.csv", na = "-99999") %>% 
  janitor::clean_names() %>% 
  uncount(count) 

```

### DATA AND METHODS ###

Data and methods will go here.

### RESULTS ### 

A) Explore the changes in annual lobster abundance (counts) by site. 

```{r, include = FALSE}
count_datula <- lob_data %>% 
  group_by(year, site) %>% 
  count(site)
  
```

```{r}
ggplot(data = count_datula, aes(x=year, y=n)) +
  geom_line(aes(color = site),
            size = 1.3,
            alpha=0.7)+
  labs(x = "Year",
       y = "Annual Count",
       color = "LTER Site",
       caption = "From SBC LTER Data, shows change in lobster size from 2012-2018") +
  ggtitle(label="Figure 1",
          subtitle = "Lobster Count from LTER Sites in Santa Barbara Channel")+
  theme(plot.title = element_text(hjust = 0.5,color = "gray2", size= 14),
        plot.subtitle = element_text(hjust = 0.5,color = "gray8", size= 8),
  ) +
  scale_x_continuous(lim=c(2012,2019), expand = c(0,0)) +
  scale_y_continuous(lim=c(0,1000), expand = c(0,0)) +
  theme_minimal()



```

B) This section will be looking at size distribution shifts of lobsters from 2012 - 2018. 
```{r}
size_data <- lob_data %>% 
  select(year,site,size_mm) %>% 
  filter(year == c("2012","2018")) %>% 
  group_by(year,site)%>% 
  summarize(
    average = mean(size_mm, na.rm = TRUE),
    sd = sd(size_mm, na.rm = TRUE)
  )

```


```{r}

ggplot(size_data, aes(x = year, y = average)) +
  geom_point(aes(color=site),
             shape=15,
             size=2.5,
             alpha=0.8) +
  labs(x = "Year",
       y = "Average Size (mm)",
       color = "LTER Site",
       caption = "From SBC LTER Data, shows change in lobster size from 2012-2018") +
  ggtitle(label="Figure 2",
          subtitle = "Lobster Size Change by Site, 2012-2018")+
  theme(plot.title = element_text(hjust = 0.5,color = "gray2", size= 14),
        plot.subtitle = element_text(hjust = 0.5,color = "gray8", size= 8),
  ) +
  scale_x_continuous(lim=c(2011,2019), expand = c(0,0)) +
  scale_y_continuous(lim=c(63,85), expand = c(0,0)) +
  theme_minimal() 


```

C) This section will compare the lobster sizes at Marine Protected Areas (MPA) vs. non-Marine Protected Areas. Naples and Isles Vista are MPAs while Arroyo Quemado, Mohawk, and Carpinteria are not. 

```{r, include = FALSE}
#First need to join the Naples (NAPL) and Isla Vista (IVEE) columns are MPAs. 

mpa_df <- lob_data %>% 
  mutate(recode_factor(site, IVEE = "MPA", NAPL = "MPA", 
                       CARP = "non-MPA", MOHK= "non-MPA",
                       AQUE = "non-MPA")
  )

mpa_clean <- mpa_df %>% 
  rename(mpa_status = `recode_factor(...)`) %>% 
  select(mpa_status, site, year, size_mm) %>% 
  mutate(year = as.factor(year),
         mpa_status = as.character(mpa_status)
  )

mpa_df_avg <- mpa_clean %>%  
  filter(year == c("2012","2018")) %>% 
  group_by(year, mpa_status) %>% 
  summarize(
    average = mean(size_mm, na.rm = TRUE),
    sd = sd(size_mm, na.rm = TRUE)
  )

mpa_nonavg <- mpa_clean %>% 
  select(mpa_status, site, year, size_mm) %>% 
  filter(year == c("2012","2018")) %>% 
  group_by(year, mpa_status)

 

#Good, looks exactly how we want it
```

```{r}
ggplot(mpa_df_avg, aes(x=year, y=average)) +
  geom_col(aes(color=mpa_status,
               fill=mpa_status),
           position="dodge2") +
  labs(x = "Year",
       y = "Average Size (mm)",
       caption = "From SBC LTER Data, shows change in lobster size from 2012-2018") +
  ggtitle(label="Figure 3",
          subtitle = "Size Changes from 2012-2018 based on MPA Status")+
  theme(plot.title = element_text(hjust = 0.5,color = "gray2", size= 14),
        plot.subtitle = element_text(hjust = 0.5,color = "gray8", size= 8),
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("darkolivegreen","dodgerblue4"),
                    name = "MPA Status")
```

```{r, include = FALSE}
#In this section we will code for the two sample t-test that will examine statistical significance. 
mpa_sample12 <- mpa_clean %>% 
  filter(mpa_status == "MPA",
         year == "2012") %>% 
  pull(size_mm)

nonmpa_sample12 <- mpa_clean %>% 
  filter(mpa_status=="non-MPA",
         year == "2012") %>% 
  pull(size_mm)
  
mpa_ttest12 <- t.test(mpa_sample12, nonmpa_sample12)

mpa_sample18 <- mpa_clean %>% 
  filter(mpa_status == "MPA",
         year == "2018") %>% 
  pull(size_mm)

nonmpa_sample18 <- mpa_clean %>% 
  filter(mpa_status=="non-MPA",
         year == "2018") %>% 
  pull(size_mm)
  
mpa_ttest18 <- t.test(mpa_sample18, nonmpa_sample18)

mpa_ttest <- t.test(mpa_sample12, mpa_sample18) 
nonmpa_ttest <- t.test(nonmpa_sample12, nonmpa_sample18)


mpa_ttest12
mpa_ttest18
mpa_ttest
nonmpa_ttest


```

```{r}

datatable(mpa_df_avg)



```


### 4) Summary







