---
title: "assignment_4_peter_omasta"
author: "Peter Omasta"
date: "11/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


### INTRODUCTION ###

Inside of this report, data collected from the Santa Barbara Channel Long Term Ecological Restoration Network (to be referred to as SBC LTER henceforth) is being used to look at how lobster populations have faired since the implementation of marine protected areas. [1] The California Spiny Lobster (*Panulirus interruptus*) is considered one of the most important valuable aquatic species from an economic perspective, with an average annual price of $ 11.07 per pound, and earning over $3.3 million in earnings for the port of Santa Barbara in 2013. [3] Over harvesting of lobsters poses a threat to the stability of the ecosystem as well as the socioeconomic systems that have been created around the harvest of lobsters in Santa Barbara. [2] Marine Protected Areas prohibit the harvesting of lobsters within the boundary, and are associated with a resulting increase in the mean size and population density of losbters. [4] Research has also been performed that examines the importantance of spillover from marine protected areas into adjacent non-protected areas. It is theorized that spillover from protected areas which have larger populations may help to buoy the populations in non-protected areas by having adult lobsters move from the densely populated protected-habitats to less dense habitats in search of food.[2] This study will look at how the designation of two sites in the Santa Barbara Channel (Isla Vista and Naples) as mrine protected areas influenced the average size and population of the California Spiny Lobster from 2012 to 2018. 

```{r, include=FALSE}
library(tidyverse)
library(janitor)
library(tidyr)
library(plotly)
library(DT)
```

```{r, include = FALSE}
#We will start by importing the necessary .csv file 

lob_data <- read_csv("lobster_abundance_sbc_lter.csv", na = "-99999") %>% 
  janitor::clean_names() %>% 
  uncount(count) 

```

### DATA AND METHODS ###

Data was collected from the SBC LTER from 2012 - 2018, as part of an on going study looking at multiple different ecological effects inside of the channel. This data specifically looks at the collection and measurement of the California Spiny Lobster from five different study sites within the SBC LTER. Measurements included the size (in milimeters) of lobsters and population counts annually from 2012 until 2018. All of the analysis was performed in R (version 3.6.1), using the tidyverse, janitor, tidyr, plotly, and DT packages. Statistical signifficance for mean body size between marine protected and non-marine protected areas was calculated using Welch's t-test with significance set at \alpha of 0.05. 

### RESULTS ### 

A) Explore the changes in annual lobster abundance (counts) by site. 

```{r, include = FALSE}
count_datula <- lob_data %>% 
  group_by(year, site) %>% 
  count(site)
  
```

```{r}
ggplot(data = count_datula, aes(x=year, y=n)) +
  geom_line(aes(color = site),
            size = 1.3,
            alpha=0.7)+
  labs(x = "Year",
       y = "Annual Count",
       color = "LTER Site",
       caption = "From SBC LTER Data, shows change in lobster size from 2012-2018") +
  ggtitle(label="Figure 1",
          subtitle = "Lobster Count from LTER Sites in Santa Barbara Channel")+
  theme(plot.title = element_text(hjust = 0.5,color = "gray2", size= 14),
        plot.subtitle = element_text(hjust = 0.5,color = "gray8", size= 8),
  ) +
  scale_x_continuous(lim=c(2012,2019), expand = c(0,0)) +
  scale_y_continuous(lim=c(0,1000), expand = c(0,0)) +
  theme_minimal() 
 



```

B) This section will be looking at size distribution shifts of lobsters from 2012 - 2018. 
```{r}
size_data <- lob_data %>% 
  select(year,site,size_mm) %>% 
  filter(year == c("2012","2018")) %>% 
  group_by(year,site)%>% 
  summarize(
    average = mean(size_mm, na.rm = TRUE),
    sd = sd(size_mm, na.rm = TRUE)
  ) 
  
  


```


```{r}

ggplot(size_data, aes(x = year, y = average)) +
  geom_point(aes(color=site),
             shape=15,
             size=2.5,
             alpha=0.8) +
  labs(x = "Year",
       y = "Average Size (mm)",
       color = "LTER Site",
       caption = "From SBC LTER Data, shows change in lobster size from 2012-2018") +
  ggtitle(label="Figure 2",
          subtitle = "Lobster Size Change by Site, 2012-2018")+
  theme(plot.title = element_text(hjust = 0.5,color = "gray2", size= 14),
        plot.subtitle = element_text(hjust = 0.5,color = "gray8", size= 8),
  ) +
  scale_x_continuous(lim=c(2011,2019), expand = c(0,0)) +
  scale_y_continuous(lim=c(63,85), expand = c(0,0)) +
  theme_minimal() 


```

C) This section will compare the lobster sizes at Marine Protected Areas (MPA) vs. non-Marine Protected Areas. Naples and Isles Vista are MPAs while Arroyo Quemado, Mohawk, and Carpinteria are not. 

```{r, include = FALSE}
#First need to join the Naples (NAPL) and Isla Vista (IVEE) columns are MPAs. 

mpa_df <- lob_data %>% 
  mutate(recode_factor(site, IVEE = "MPA", NAPL = "MPA", 
                       CARP = "non-MPA", MOHK= "non-MPA",
                       AQUE = "non-MPA")
  )

mpa_clean <- mpa_df %>% 
  rename(mpa_status = `recode_factor(...)`) %>% 
  select(mpa_status, site, year, size_mm) %>% 
  mutate(year = as.factor(year),
         mpa_status = as.character(mpa_status)
  )

mpa_df_avg <- mpa_clean %>%  
  filter(year == c("2012","2018")) %>% 
  group_by(year, mpa_status) %>% 
  summarize(
    average = mean(size_mm, na.rm = TRUE),
    sd = sd(size_mm, na.rm = TRUE)
  )

mpa_nonavg <- mpa_clean %>% 
  select(mpa_status, site, year, size_mm) %>% 
  filter(year == c("2012","2018")) %>% 
  group_by(year, mpa_status)

 

#Good, looks exactly how we want it
```

```{r}
ggplot(mpa_df_avg, aes(x=year, y=average)) +
  geom_col(aes(color=mpa_status,
               fill=mpa_status),
           position="dodge2") +
  labs(x = "Year",
       y = "Average Size (mm)",
       caption = "From SBC LTER Data, shows change in lobster size from 2012-2018") +
  ggtitle(label="Figure 3",
          subtitle = "Size Changes from 2012-2018 based on MPA Status")+
  theme(plot.title = element_text(hjust = 0.5,color = "gray2", size= 14),
        plot.subtitle = element_text(hjust = 0.5,color = "gray8", size= 8),
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("darkolivegreen","dodgerblue4"),
                    name = "MPA Status")
```

```{r, include = FALSE}
#In this section we will code for the two sample t-test that will examine statistical significance. 
mpa_sample12 <- mpa_clean %>% 
  filter(mpa_status == "MPA",
         year == "2012") %>% 
  pull(size_mm)

nonmpa_sample12 <- mpa_clean %>% 
  filter(mpa_status=="non-MPA",
         year == "2012") %>% 
  pull(size_mm)
  
mpa_ttest12 <- t.test(mpa_sample12, nonmpa_sample12)

mpa_sample18 <- mpa_clean %>% 
  filter(mpa_status == "MPA",
         year == "2018") %>% 
  pull(size_mm)

nonmpa_sample18 <- mpa_clean %>% 
  filter(mpa_status=="non-MPA",
         year == "2018") %>% 
  pull(size_mm)
  
mpa_ttest18 <- t.test(mpa_sample18, nonmpa_sample18)

mpa_ttest <- t.test(mpa_sample12, mpa_sample18) 
nonmpa_ttest <- t.test(nonmpa_sample12, nonmpa_sample18)


mpa_ttest12
mpa_ttest18
mpa_ttest
nonmpa_ttest


```

```{r}

datatable(mpa_df_avg)

```


### Summary

Marine Proected Areas like the Isla Vista (IVEE) and Naples (NAPL) sites saw an increase in the number of lobsters from 2012 to 2018 (figure 1) as well a increase in the mean size (~12mm) of lobsters (figures 3,4) in the same period of time. This result was statistically significant, and it may imply that protecting the areas allowed for populations to mature and reproduce which resulted in greater numbers and lobster size. On the other hand, non-Marine Protected Sites saw either static (Carp) and (Aque) or a decrease (MOHK) in the number of lobsters found from 2012 to 2018 (figure 2) There was no statistically significant change in average size between the two samples during the same time period. 

Designating an area as a Marine Protected Area seems to have a demonstrably positive effect on lobster populations and size. This increase in mean size and overall population may have also resulted in a spillover to the other non-marine protected areas which is why they did not have as much of a decrease in population size and mean body size as might be expected with harvesting activites.Keeping an area as a non-Marine Protected Area does not have any statistically negative effects on the populations, but decreases in mean population size, and the overall number of individuals may signal an early trend to system managers about over harvesting of lobsters and the need to rotate the Marine Protected Area designation to further conserve lobsters for ecosystem and socioeconomic purposes. 

### Citations

1) Reed D. 2019. SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012. Environmental Data Initiative. https://doi.org/10.6073/pasta/a593a675d644fdefb736750b291579a0. Dataset accessed 10/30/2019.

2) Kay, Matthew C., Hunter S. Lenihan, Carla M. Guenther, Jono R. Wilson, Christopher J. Miller, and Samuel W. Shrout. “Collaborative Assessment of California Spiny Lobster Population and Fishery Responses to a Marine Reserve Network.” Ecological Applications 22, no. 1 (2012): 322–35. https://doi.org/10.1890/11-0155.1.

3) News, Fishermen’s. “Commercial Fishing’s Contribution to Santa Barbara Economy.” Fishermen’s News. Accessed November 13, 2019. https://www.fishermensnews.com/story/2015/12/01/features/commercial-fishings-contribution-to-santa-barbara-economy/364.html.

4) Lester, Sarah E., Benjamin S. Halpern, Kirsten Grorud-Colvert, Jane Lubchenco, Benjamin I. Ruttenberg, Steven D. Gaines, Satie Airamé, and Robert R. Warner. “Biological Effects within No-Take Marine Reserves: A Global Synthesis.” Marine Ecology Progress Series 384 (May 29, 2009): 33–46. https://doi.org/10.3354/meps08029.


